---
title: "week6 hip replacement"
author: "Melanie Rees-Roberts"
date: "2024-11-07"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 6 practical: Hip replacement data work

#This is work to complete the Hip replacement data practical work for IntroHDS 
#course, week 6. 
#The following questions require answering to complete the practical:

#1. Plot 'EQ-5D Index' scores pre and post operation for each gender

#2. Calculate how many patients in this dataset have been told by a doctor that 
#they have problems caused by a stroke

#3. Create a clean and tidy table with pre and post operation activity levels

## Libraries and packages needed 

#For this work we require the following packages; tidyverse

#From this package we will use the following libraries

```{r call libraries}
library(readr)
library(tidyr)
library(dplyr)
library(here)
library(ggplot2)
```

## Import the data
#Data file provided called Hip replacement CCG 1819. Import the data with 
#read_csv command

```{r}
hip <- read_csv("Hip Replacement CCG 1819.csv")
```
## Understand, select and tidy the data

```{r}
head(hip)
colnames(hip)

#Check this is for a single year - then dont need the year variable
hip %>% 
count(Year)
```

#For this exercise, we need the following data variables:
#Gender, Strike, Pre-OP Q EQ5D Index, Post-Op Q EQ5D Index, Pre-Op Q Activity, Post-Op Q Activity
#Create tibble with required variables, shortening the variable names where appropriate

```{r}
hip2 <- hip %>% 
  select('Gender','Pre-Op Q EQ5D Index','Post-Op Q EQ5D Index','Stroke','Pre-Op Q Activity','Post-Op Q Activity') %>% 
  rename(EQ5D_pre = 'Pre-Op Q EQ5D Index',
         EQ5D_post = 'Post-Op Q EQ5D Index',
         activity_pre = 'Pre-Op Q Activity',
         activity_post = 'Post-Op Q Activity')

head(hip2)

```

#Remove missing values - create this into new data frame hip3

```{r}
hip2$Gender %>% table()

hip3 <- hip2 %>% 
  drop_na() %>% 
  filter(Gender !='*')

head(hip3)

```

##Tidy the data - move pre and post into a time variable and th EQ5D and Activity into variables
#Saved into a new dataset tidy_hip

```{r}
tidy_hip <- hip3 %>% 
  pivot_longer(c(EQ5D_pre,EQ5D_post),
               names_to = 'Time_EQ5D',
               names_prefix = 'EQ5D_',
               values_to = 'EQ5D')


tidy_hip <- tidy_hip %>% 
  pivot_longer(c(activity_pre, activity_post),
               names_to = 'Time_activity',
               names_prefix = 'activity_',
               values_to = 'activity')


head(tidy_hip)
```

##Answer the questions

##1. Plot 'EQ-5D Index' scores pre and post operation for each gender

```{r}
#From tidy_hip we can now select the data and plot the scores

tidy_hip_plot <- tidy_hip %>% 
  select(EQ5D, Time_EQ5D,Gender
         )

head(tidy_hip_plot)

#Now plot the data pre and post for each gender

tidy_hip_plot %>% 
  ggplot() +
  geom_boxplot(aes(x=Time_EQ5D, y= EQ5D, colour = Gender))
```


##2. Calculate how many patients in this dataset have been told by a doctor that 
#they have problems caused by a stroke

```{r}
#We need to return to our table where we did not make the data tidy - this may cause duplication of Stroke data
#calculate the how many patients have been told they have problems caused by stroke

hip2$Stroke %>% table() 
```
# A total of 400 patients in this dataset have been told that they have problems caused by stroke.

##3. Create a clean and tidy table with pre and post operation activity levels
#We converted the pre- and post- activity levels earlier in the file - see code above. However, to show just these variables

```{r}

head(tidy_hip)

tidy_hip_activity <- tidy_hip %>% 
  select(Time_activity,activity) 

head(tidy_hip_activity)
```

